BExIS Manual for Traitdata

### 2018-02-13 bexis_handbook/bexis_handbook_traits.Rmd
add BExIS handbook draft

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
- name: Malte Jochum \footnotemark[4]
- name: Ga\"etane Le Provost \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: Markus Fischer
- name: Pete Manning 
- name: Birgitta König-Ries
- name: Nadja K. Simons \footnotemark[6]
csl: amnat.csl
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F), Senckenberganlage 25, 60325 Frankfurt am Main, Germany}
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[5]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[6]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}

\linenumbers

# Abstract

1.	The use of trait-based approaches is now widespread in all domains of ecological research, thanks to the great potential for such data to deliver general conclusions regarding the nature of biodiversity. Accordingly, a wealth of trait data is now available for many organism groups but due to lacking standardisation these data come in heterogeneous data formats.
2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
3. In order to facilitate the standardisation and harmonization of distributed trait datasets, we propose a general and simple vocabulary and a two-dimensional data structure for storing and sharing ecological trait data. 
4. We provide an R-package that enables the transformation of any data into the proposed form. This also allows to harmonize and merge trait datasets from heterogeneous sources, which facilitates the compilation of databses of particular research focus. 
5. With these decentralized tools for trait-data harmonization, we intend to facilitate the exchange and analysis of trait data for ecological research and enable a global synthesis of traits across a wide range of organisms and ecosystems.

# Key-words:

functional ecology, functional traits, semantic web, ontologies, data standard

# Introduction

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07]. Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
For example, trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity and ecosystem functioning (refs). 
Using traits is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring both the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.
With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15].

<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. <!-- GLP: Maybe not the best example as they are not looking at trait diversty and species interactions ? 
Put instead:
Gravel et al. 2016 The meaning of functional trait composition of food webs for ecosystem functioning
And Coux et al. 2016 Linking species functional roles to their network roles
?
-->
<!--
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @simons16; @gossner16; @moretti13;  @kuhsel15;  @birkhofer14], as well as macro-patterns of biodiversity at the global scale [@diaz16; e.g. @reich97]. <!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].--> <!-- GLP: What do you mean by processes ? I’m not sure that the literature cited is always appropriate here ?
-->
<!--
Most importantly, the trait framework relates key ecosystem services for human well-being [@diaz11; @lavorel12; @allan15] back to changes in biodiversity in consequence to global change or local anthropogenic land use [ @diaz07; @gossner16; @mangels17; @jennings09; @simons16;  @mouillot13;  @vandewalle10;  @birkhofer15;@gamez-virues15; @birkhofer17]. 
--> 
<!-- CP: I like very much this whole paragraph (both writing and content), but I feel that it’s a bit out of the topic. It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important). I give some suggestions on how to cut it and rather develop more the following one, which people are surely less aware of and which sets more the aim of the paper. I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.

That’s the structure I try to follow:
1. Traits are important in ecology, evolution and conservation (i.e. this paragraph)

2. To increase the scale of studies (spatial, temporal, taxonomical) we need to put together trait data from different sources. When we do this we can answer new questions and find generalities. So there is a growing interest in assembling trait data from different sources and taxa (e.g. give the number of papers using TRY).

3. There is a problem of standardization that makes it difficult to match trait datasets from different sources. Different reasons: no traits definitions, no common structure.

4. One solution are databases. But these are costly and depend on the will of few persons/funding. They also cover few taxonomic groups. Matching between different databases is still difficult because of different definitions and structures. Most data are still uploaded separately.
We need a solution to make their use easy.

But of course it’s just a suggestion, feel free to happily ignore it!

I would choose 2 or 3 (also for the previous), anyway it’s impossible to be exhaustive. Maybe put (e.g. ref1, ref2) so make clear that it’s just examples

-->
<!-- GLP: 
The second paragraph is a bit confusing for me.

I wonder also if the introduction could be restructures like this :

Trait based approaches and their use
From the individual to population
Community level (assemblage, interactions, effects on ecosystem)
Macroecological question

 Add a general paragraph on the fact that there have been some initiatives for the development of trait data base for plants (e.g. TRY), why is it important (provide quick access to plant trait values, allowing comparisons even between ecosystems and biomes), now the development of trait data base for other organisms (from microbes to mammals). Say also something on the development of general protocols for measuring traits (e.g. Moretti et al. 2016; Freschet et al. 2017) ?

Importance of data standardization. This is the main idea here and so I wonder if some idea of the first point (Need for data standardization) should not be merge here with the last paragraph of the introduction ? As there is also some repetitions between those 2 parts.

There is just some copy-past to do. I didn’t change the structure of the introduction but if you like the structure that I proposed, we can discuss next week or I can help you with that. I like to write introduction 

I wonder if you should speak about ‘Functional traits’here or just Traits ? As you are not only considered Functional traits in this paper (e.g. life history traits) ? Maybe remove the definition no ? 
-->
<!-- PM: Thuis is quite complex- e.g. ‘taxonomic impediment’etc. I actually think you can simplify this and built these references into the more general text above by adding a copule of more examples. I don’t think you need to say much about the merits of the trait based approach here- it’s covered elsewhere.

Huge number of references here- I think you can strip it down to a few key ones, and reviews. Overall I think this section is quite a lot more complex than it needs to be – I think all you need to say is that it’s very popular, general and mechanistic and so adopted throughout eco and evo, with no sign of slowing down- the TRY database download stats can provide a quantitative measure of this- see their site. 
-->
<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, which continues to integrate a rapidly increasing number of fundamental and applied research questions. 
Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->

However, fully exploiting this portential relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa. <!-- PM:  Not 100% clear what you mean here- factual information could be anything- can you be more specific? -->
To mobilize data for this enterprise <!-- PM: Please define what the exercise is- too vague at present
-->, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources. <!-- PM: I think this is a key justification- perhaps say what happens if we don’t do this – non repeatable work, differing definitions of traits that may lead to scientific flaws, repetition of the painstaking task of processing trait data e.g. to compute species level means. 
-->
The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way. <!-- CP: I felt that this needed a definition, not sure everybody know what an ontology is. Maybe we should guide more the reader that might not be familiar with the problems.
-->

Nonetheless, individual researchers often neglect the benefit of centralised databases <!-- PM: Explain what this benefit is. -->. Furthermore, many journals and research funding agencies require  data to be published without access restrictions [@swan12; @allison15]. However, this primary data is often posted on file hosting services, where no data standards are forced upon the uploaded material (e.g. Figshare, Dryad). The long-term outcome of this situation is an abundance of freely available data, but a lack of data standardisation.

In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from a pragmatic view of data providers and data users. We further (2.) review existing trait databases and web services, as well as initiatives for standardisation. 
We (3.) describe the common practice <!--PM: Current or your own new definition?--> for structuring trait data for ecological research. 
Based on these considerations, we (4.) propose a low-threshold <!-- PM: What’s this? Please define
-->  structure and vocabulary for describing trait-datasets compatible with existing ontologies <!-- PM & CAT: define--> . (5.) Finally, we provide an R-package that assists in aply this structure to own data and for the harmonisation of data from distributed sources.

With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards simple and low-threshold <!-- PM:Again, I’m not sure what this is a low threshold of? --> trait data standardisation.

Despite the obvious benefit and great potential of trait-based approaches, integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used. All of these factors render trait data extremely heterogeneous.

As a response to these issues, trait data have been standardized in centralized databases for specific organism groups and research questions.
<!-- PM: (refs, also give an e.g. or two for some big ones like TRY- saw there was a new one for tropical mammals a couple of days ago too and a huge Chinese plants one- in )--> 
These compiled data are either being published as static data tables or in the form of sophisticated online portals like the plant trait database TRY (see below for a review of current initiatives). These portals attract data submissions from a focal community of researchers and map heterogeneous data into a common scheme, and importantly also, provide networks for consensus building in measurement methods. Furthermore, database services offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets.

In contrast to standardized databases, open-access data publication in low-threshold repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org is becoming increasingly frequent. Such repositories foster decentralised data hosting with low thresholds regarding data standardisation and documentation.

Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets.
To cope with this trend towards distributed data sources, data brokering services (e.g. the German Federation on Biological Data; gfbio.org)  or database <!-- AO: Just “data” or “research data”--> management systems for scientific projects (e.g. BEXIS, XX <!--BKR: Diversity Workbench (developed at SNSB), Metacat (https://knb.ecoinformatics.org/)--> ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.

However, to mobilize the broad wealth of distributed data into structured databases <!-- AO: Maybe better: formats-->,  a certain minimum of information from original data providers is requrired, e.g. on the methodology and definitions of trait values or the covariates describing the ecological context of the trait observation. In order to get this accurate and well-defined input ecological researchers need to be aware of and understand basic principles of data standards and ontologies.  <!-- AO: The last paragraphs may need more sources/proofings?--><!-- PM: Literacy is quite a strong word- as illiterate is quite an insult. I think awareness is a problem too.  -->

 More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.

# 1. A review of trait-standard initiatives

In this section, we review four kinds of initiatives that are of relevance for trait data standardisation: 
the first are projects that work with trait datasets that have been assembled out of a particular research interest, either by measurement or colated from literature; 
the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
the third aim at the standardization and develops consensus methods and definitions for traits by providing classifications and thesauri; 
the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct 'ontologies' for an application in a 'semantic web' of information.

We discuss these initiatives seperately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the TOP Thesaurus of plant traits [@garnier17]. 
We show how the degree of trait-data standardization in exising datasets spans along this entire spectrum and which tools and standards do exist to acheive harmonization of data from multiple, distributed sources. 
The objective of this review is to raise awareness for the generic structure of trait data and aid ecological researchers to share and publish own datasets in the appropriate form.

## Trait datasets

In the field of comparative physiology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been contained in regular articles or books. 
With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13;  @ricklefs17].

Today, the online publication of such data is greatly facilitated by hosting services like  Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org. 
Furthermore, these services warrant long-term accessibility (offering DOIs for referencing), and support Public Domain dedication or Creative Commons Attribution licenses, as increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12]. <!-- AG: Perhaps mention the Bouchout declaration here: http://www.bouchoutdeclaration.org/ ? -->
However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk.

Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds in terms of metadata documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources on data management. 
In most of these datasets, the column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives. Datasets vary in the detail provided in the metadata, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].

As it stands, this lack of data standardization of low-threshold online repositories will mean that that the compilation of data into larger collections will be inefficient and that many published datasets will not be properly exploited.

A global overview of existing trait data for all taxa and trait types is difficult to obtain. Therefore, in an attempt to collate a list of existing, distributed and unstandardized datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy; Appendix A) which lists published trait datasets, their regional and taxonomic focus, the traits covered, their location on the internet and the terms of their use (see Appendix 1 for a current excerpt of this list). We invite data owners and users to register further datasets qualifying as trait data to this spreadsheet.

## Database initiatives

<!-- community database backend, but static publication --> 
In the past two decades,  distributed trait data have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g. @klotz02; @iversen17; @kleyer08; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09]. <!--[@wright04; ] -->
While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
Just as the individual datasets described above, many of these databases are published as static datasets on low-threshold file hosting platforms and are updated irregularly.

Dealing with much larger amounts of data, initiatives that form around museal collection data are more concerned with data standardization. Aiming for a mobilization of measurements from collections, the  VertNet database compiles and harmonized large quantities of vertebrate traitdata [@guralnick16]. The resulting data are published as static datasets which are updated as new data sources are integrated.

<!-- database backend, selective download, user access management --> 
More specialised trait-database platforms have been created that cover certain trait types (e.g. floral traits, seed traits or wood density traits), interaction types (e.g. pollination traits or feeding relationships), or environmental and experimental context of the trait observation (e.g. location or climatic data).  
These centralized database initiatives attract data submissions from a defined research field and take care of the harmonization process and greatly facilitate data synthesis. 
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
By aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other platform. The online platform enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
On the side of animals, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/). 
Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is still lacking. The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans make harmonizing animal trait data a more complex task than for plant traits [@moretti17].

<!--Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12]. Here, taxa are derived from metagenomic analysis [@torsvik02; @langille13]. Databases are also used to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g. the KEGG orthology, @kanehisa12]. -->

Regarding open access, hardly any of these centralized databases for functional traits comply with the criteria demanded by journals and funding agencies for primary data publication. They incentivize data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.  
Time will show how these incentives fare with an increasing availability of open access trait data  on low-threshold data repositories. 
<!-- AO: In a sense of data availability to the public : yes. But do we discuss it here? We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs. non-open data, or?
bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
 -->

## Thesauri and Ontologies for traits

A major problem in trait-data standardisation is the reference to accepted and unambiguous trait definitions. This is acheived via thesauri and ontologies. 
In database management, a thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].  Expanding on this, ontologies link the defined terms with each other by formally defining the relationships between them, with the objective to enable a computational interpretation of data. Providing relational ontologies in a machine readable form makes them compliant with semantic web standards, thus supporting the goal of widely interlinked data structures for the world wide web [@berners-lee01; @page08].

<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. --><!-- BKR: Isn’t that the same? Or can you have multiple parents for one node? FDS: agree. simplify phrasing. -->

Standard trait definitions have been provided in all shades along this continuum, ranging from listings of definitions in glossaries, over well-defined thesauri to relational ontologies. The initiatives behind these are first and foremost concerned with consensus building on trait definitions. 
Assigning an unambiguous methodological protocol of measurements to a trait is key for harmonizing the physical process of measuring. Efforts to develop handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but are of limited effect when harmonizing trait data that predate or ignore this standard [@kattge11a].
Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
Thesauri of trait definitions have been developed for plants [e.g. the Planteome ontology for crop plants, @walls12; @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g. the Hymenoptera Anatomy Ontology, @yoder10; the vertebrate trait ontology, @park13]. 
Comprehensive trait ontologies, including hierarchies of parent terms, have been developed in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], and on the side of animal traits, in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].

These existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. 
For biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access.  <!-- Mention also NCBO BioPortal? https://bioportal.bioontology.org/ -->
These publicly available online resources also enable linking trait measurements to globally unique Uniform Resource Identifiers (URIs) when referencing traits from within own datasets. For example, a measurement of seed size would be linked to Planteome Trait Ontology definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'.

To be unambiguous, any thesaurus should be defining terms based on other well-defined terms from semantic ontologies. The definition provided here contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
This way, each trait definition may link to a broader or narrower term. For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'. 
Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies. By providing this semantic interlinkage of trait ontologies, a web of definitions is spun across the internet which allows researchers and search engines to relate independent trait measurements with each other.

The benefit of such ontologies will increase if webservices and open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. To harmonize trait data across databases, future trait terminology initiatives should provide this functionality.
More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from different sources.

## Trait-data structures for the semantic web

While trait thesauri define traits for focal groups of organisms, they are not specifying the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g. for taxa. 
To make sense of trait data in the context of more general databases, a consensus definition of trait datasets is inevitable.

Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative measurements or qualitative facts (i.e. trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
Expaning on this definition, quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig. 1*a*), by aggregating replicated measurements on multiple entities (Fig. 1*b*) or by estimation the means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).

\begin{figure}[t]
\centering
\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
\caption{Types of ecological trait data assume different entities or reported quantities. a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g. reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g. a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire taxon.}

\end{figure}
<!-- DF : The “variables” for the taxa etc should only be reused for the different examples when they refer to something similar: “a) Entity: an individual X of taxon Y”, “b) Entity: taxon Y”, “c) Entity: males of taxon Z” and “d) Entity: genus U”. Alternatively, since these are already example, why use variables here in the first place and not just specific example for the taxa as well: “a) Entity: an individual X of taxon Tettigonia viridissima” -->

These relations between a trait observation and an individual specimen as an occurrence of a particular taxon have been formalized in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12]. 
These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining which field is to contain trait values, units, taxon names, variables such as sex or life stage, as well as ancillary information of time and date of observation, as well as methodological details. The terminologies provided by these standards are quite universal and cover most use cases of trait data. 
<!-- TODO FDS: dimensional and star-shema databases, relational databases Kattge et al 2011, Kimball & Ross 2002-->
<!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter. The Trait Data Standard is a standard on how to store trait data measurements. In this chapter you mention what databases for trait data there are (usually specific for certain taxa) and what ontologies there for the different traits, however it is missing the section on how trait data measurements have been and can be stored prior to your standard. For instance the ABCD standard has a huge section for MeasurementsAndFacts. And while it certainly is not primarily made to capture trait measurements, it can certainly be used for it (at least on a unit level, combining measurements for higher taxa is not possible). Darwin Core has similar, though not as extensive, fields for measurements, as they were modelled based on ABCD. While I agree that your standard is useful and necessary, it certainly is not the first standardized way to describe trait measurements.  -->

Specifically designed for the context of plant traits, @kattge11 proposed a generic structure for plant trait data, that covers most potential use cases. This data structure is build around a central data table that contains observations, i.e. a single event of measurements on the same individual plant specimen at the same point in time. Identifiers link multiple measurements to the same observation, each measurement being well defined by additional standard tables. The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment. This structure can be implemented in any software or programming language for relational databases. 
<!-- FDS: enter comments of Jens Kattge here!-->

In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives]. For example, the Common Daisy (*Bellis perennis*) would be related via a globally unique taxonID using the URI https://www.gbif.org/species/3117424 from the GBIF Backbone Terminology. The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review. 
Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/]. 
The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11].

However, these proposed standards and the available terms and extensions are responses to a demand from biodiversity data management for more structured input from the research community. 
None of the aforementioned aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards. 
One reason for this is most certainly complexity: The data structures are designed for multi-layered databases rather than for static datasets for which a two-dimensional data table might suffice.  
If the traits are assembled for a specific research question with little ancillary information, it may be safer to add them as extra columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road. This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet. 
<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
The trade-off is data-consistency and readability *vs.* avoidance of content duplication.

<!--When traits are assessed on a higher level of organisation, e.g. microbial functional traits assessed at the community level taken from a soil sample, key data columns would be placed in the same table as the core data are traits. Here, location or measurement information are the primary focus of the investigation and must be contained in the . 
<!-- PM: I get lost here-->
<!--A general recommendation, whether a column describes ancillary data or is part of the central dataset is ill advised. Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context. -->
<!-- FDS: shorten this and refer to existing standards for database use. the star-structure of ABCD, Kattge et al , and Garnier .-->

<!--while the existing trait data standards may cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots; PM: Vague- not sure what is missing here- be more specific --> 
The other reason is lack of awareness and literacy: many trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off, which is not affordable for small and intermediate size research projects.
By filling this gap, data brokering services (e.g. the German Federation on for Biological Data; gfbio.org) or database management systems for large-scale scientific projects (e.g. BExIS, NEON ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into consistent and interlinked data infrastructures. One goal of such initiatives is to facilitate the publication and standardization of data for researchers, e.g. by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.<!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->

>>>>

---
title: "BExIS Manual for Traitdata"
author: "Florian D. Schneider"
date: "13 February 2018"
output:  pdf_document
bibliography: ../manuscript/div05.bib
csl: ../manuscript/amnat.csl
---

Functional traits have become a core concept of ecological research and have been assessed for many organism groups within the Biodiversity Exploratories project.

Sharing and harmonizing trait data into databases is a key task of biodiversity data management, but it relies on an unambiguous assignment of trait definitions and taxa, as well as clearly defined datasets. To facilitate the data handling by BExIS and other downstream services (e.g. GFBio or TRY) we defined a standard vocabulary for publishing trait datasets.

This document describes the main features of the standard vocabulary and how to prepare traitdata for upload into BExIS.

# Traitdata vocabulary

The full set of terms can be accessed at https://ecologicaltraitdata.github.io/ETS/.

The vocabulary proposed here is meant for two-dimensional spreadsheet data. These simple table formats are what a typical research project on a particular organism group and locality produces. The vocabulary includes a core section, as well as a series of terms which are used to define the level of the reported measurement or fact (e.g. measured on individuals, or reported for an entire species), or to relate it to a broader experimental or ecosystem context. The vocabulary also contains terms for describing metadata acompanying the trait dataset.^[Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain. For datasets containing multi-layered information on observations, traits, taxa and environmental context, we recommend the use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr15, ideally compiled into Darwin Core Archives [DwC-A, @robertson09]. ]

We assume that data are stored in a observation table, where each row is the reported measurement or fact for a single observation. At minimum, the following columns are required: 1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 2. a descriptive trait name (`traitName`);  3. the scientific taxon for which the measurement or fact was obtained (`scientificName`). 
<!--For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. However, in order to keep the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers. -->

## Identifiers

To make data compatible with a consensus standard of taxon names or or look-up tables of traits, it is highly recommended to map them to globally valid terminologies by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively; e.g. GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass').

By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11]. This redundancy of data allows for continuity on the authors and data owners side while also ensuring quality checks and comparability for the data user.

## Extensions

Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event. Here, the assignment of a unique identifier per single measurement event (`measurementID`) or individual occurrence (`occurrenceID`), defines the structure these data would take in a relational database context.
